Connecting R with D3 for dynamic graphics, to explore multivariate data with tours

### 2018-03-02 paper-tourrd3.Rmd
Small changes only

---
title: Connecting R with D3 for dynamic graphics, to explore multivariate data with tours
author:
  - name: Michael Kipp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  mkipp271@gmail.com
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
  - name: Ursula Laa
    affiliation: Monash University
    address:
    - Department of Physics
    email:  ursula.laa@monash.edu
bibliography: references.bib
link-citations: true
fig_caption: yes
abstract: >
  The tourr package in R has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections. It can display as a movie but has no capacity for interaction, such as stop/go, change tour type, drop/add variables. The tourrGui  package provides these sorts of controls, but the interface is programmed with the dated RGtk2 package. This work explores using custom messages to pass data from R to D3  for viewing, using the Shiny framework.  
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

@@

The tour algorithm [@gt_pp; @gt_pp_mc] is a way of systematically generating and displaying projections of high-dimensional spaces in order for the viewer to examine the multivariate distribution of data. It can do this either randomly, or by picking projections judged interesting according to some criterion or index function. The tourr package [@tourr] provides the computing and display in R [@R; @ihaka:1996] to make several types of tours, grand, guided, little and local. The projection dimension can be chosen between one and the number of variables in the data. The display, though, has no capacity for interaction. The viewer can watch the tour like a movie, but not pause it and restart, or change tour type, or number of variables.

These interactive controls were provided with the tourrGui package [@tourrGui], with was programmed with the RGtk2 package [@RGtk2]. This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like Shiny [@shiny]. To display dynamic graphics though, is not straight-forward. This paper explains how to use D3 [@D3] as the display engine, in a Shiny GUI, using custom message passing between server and client.

@@

A tour path is a smooth sequence of projection matrices, $p\times d$, that when combined with data, $n\times p$, and a rendering method, produces a steady stream of $d$-dimensional views of the data.  Each tour is initialised with the `new_tour()` method, which instantiates a tour object and takes as arguments the data $X$, the tour method, e.g. `guided_tour()`, and the starting basis. Once initialised, a new target plane is chosen, and a series of steps along a geodesic path from starting to target plane are generated by interpolation.

@@

D3.js (Data-Driven Documents) (REF) is a JavaScript library for manipulating documents based on data. The advantages of D3 are similar to those provided by Shiny: namely, an industry standard with rich array of powerful, easy to use methods and widgets that can be displayed on a wide variety of devices, with a large user base. D3 works on data objects in the JavaScript Object Notation (JSON) format, which are then parsed and used to display customisable data visualisations.

@@

The projections are calculated using the tour object in an `observe()` environment, which repeatedly runs the code until a reactive variable is changed, at which point it is is "invalidated" and then re-started. The projections are calculated using the following code
block:

>>>>

---
title: Connecting R with D3 for dynamic graphics, to explore multivariate data with tours
author:
  - name: Michael Kipp
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  mkipp271@gmail.com
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
  - name: Ursula Laa
    affiliation: Monash University
    address:
    - School of Physics and Astronomy
    email:  ursula.laa@monash.edu
bibliography: references.bib
link-citations: true
fig_caption: yes
abstract: >
  The tourr package in R has several algorithms and displays for showing multivariate data as a sequence of low-dimensional projections. It can display as a movie but has no capacity for interaction, such as stop/go, change tour type, drop/add variables. The tourrGui  package provides these sorts of controls, but the interface is programmed with the dated RGtk2 package. This work explores using custom messages to pass data from R to D3  for viewing, using the Shiny framework.  
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

@@

The tour algorithm [@gt_pp; @gt_pp_mc] is a way of systematically generating and displaying projections of high-dimensional spaces in order for the viewer to examine the multivariate distribution of data. It can do this either randomly, or by picking projections judged interesting according to some criterion or index function. The tourr package [@tourr] provides the computing and display in R [@R; @ihaka:1996] to make several types of tours: grand, guided, little and local. The projection dimension can be chosen between one and the number of variables in the data. The display, though, has no capacity for interaction. The viewer can watch the tour like a movie, but not pause it and restart, or change tour type, or number of variables.

These interactive controls were provided with the tourrGui package [@tourrGui], with was programmed with the RGtk2 package [@RGtk2]. This is not the toolkit of choice today, and has been superceded with primarily web-capable tools, like Shiny [@shiny]. To display dynamic graphics though, is not straight-forward. This paper explains how to use D3 [@D3] as the display engine in a Shiny GUI, using custom message passing between server and client.

@@

A tour path is a smooth sequence of projection matrices, $p\times d$, that when combined with a matrix of n data points, $n\times p$, and a rendering method, produces a steady stream of $d$-dimensional views of the data.  Each tour is initialised with the `new_tour()` method, which instantiates a tour object and takes as arguments the data $X$, the tour method, e.g. `guided_tour()`, and the starting basis. Once initialised, a new target plane is chosen, and a series of steps along a geodesic path from starting to target plane are generated by interpolation.

@@

D3.js (Data-Driven Documents) [@D3] is a JavaScript library for manipulating documents based on data. The advantages of D3 are similar to those provided by Shiny: namely, an industry standard with rich array of powerful, easy to use methods and widgets that can be displayed on a wide variety of devices, with a large user base. D3 works on data objects in the JavaScript Object Notation (JSON) format, which are then parsed and used to display customisable data visualisations.

@@

The projections are calculated using the tour object in an `observe()` environment, which repeatedly runs the code until a reactive variable is changed, at which point it is "invalidated" and then re-started. The projections are calculated using the following code
block:

### 2018-03-02 paper-tourrd3.tex
Small changes only

The tour algorithm (Cook et al. 1995; Cook et al. 2007) is a way of
systematically generating and displaying projections of high-dimensional
spaces in order for the viewer to examine the multivariate distribution
of data. It can do this either randomly, or by picking projections
judged interesting according to some criterion or index function. The
tourr package (Wickham et al. 2011) provides the computing and display
in R (R Core Team 2018; Ihaka and Gentleman 1996) to make several types
of tours, grand, guided, little and local. The projection dimension can
be chosen between one and the number of variables in the data. The
display, though, has no capacity for interaction. The viewer can watch
the tour like a movie, but not pause it and restart, or change tour
type, or number of variables.

These interactive controls were provided with the tourrGui package
(Huang, Cook, and Wickham 2012), with was programmed with the RGtk2
package (Lawrence and {Temple Lang} 2010). This is not the toolkit of
choice today, and has been superceded with primarily web-capable tools,
like Shiny (Chang et al. 2017). To display dynamic graphics though, is
not straight-forward. This paper explains how to use D3 (Bostock,
Ogievetsky, and Heer 2011) as the display engine, in a Shiny GUI, using
custom message passing between server and client.

@@

A tour path is a smooth sequence of projection matrices, \(p\times d\),
that when combined with data, \(n\times p\), and a rendering method,
produces a steady stream of \(d\)-dimensional views of the data. Each
tour is initialised with the \texttt{new\_tour()} method, which
instantiates a tour object and takes as arguments the data \(X\), the
tour method, e.g. \texttt{guided\_tour()}, and the starting basis. Once
initialised, a new target plane is chosen, and a series of steps along a
geodesic path from starting to target plane are generated by
interpolation.

@@

D3.js (Data-Driven Documents) (REF) is a JavaScript library for
manipulating documents based on data. The advantages of D3 are similar
to those provided by Shiny: namely, an industry standard with rich array
of powerful, easy to use methods and widgets that can be displayed on a
wide variety of devices, with a large user base. D3 works on data
objects in the JavaScript Object Notation (JSON) format, which are then
parsed and used to display customisable data visualisations.

@@

The projections are calculated using the tour object in an
\texttt{observe()} environment, which repeatedly runs the code until a
reactive variable is changed, at which point it is is ``invalidated''
and then re-started. The projections are calculated using the following
code block:

@@

\address{%
Ursula Laa\\
Monash University\\
Department of Physics\\
}
\href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}

>>>>

The tour algorithm (Cook et al. 1995; Cook et al. 2007) is a way of
systematically generating and displaying projections of high-dimensional
spaces in order for the viewer to examine the multivariate distribution
of data. It can do this either randomly, or by picking projections
judged interesting according to some criterion or index function. The
tourr package (Wickham et al. 2011) provides the computing and display
in R (R Core Team 2018; Ihaka and Gentleman 1996) to make several types
of tours: grand, guided, little and local. The projection dimension can
be chosen between one and the number of variables in the data. The
display, though, has no capacity for interaction. The viewer can watch
the tour like a movie, but not pause it and restart, or change tour
type, or number of variables.

These interactive controls were provided with the tourrGui package
(Huang, Cook, and Wickham 2012), with was programmed with the RGtk2
package (Lawrence and {Temple Lang} 2010). This is not the toolkit of
choice today, and has been superceded with primarily web-capable tools,
like Shiny (Chang et al. 2017). To display dynamic graphics though, is
not straight-forward. This paper explains how to use D3 (Bostock,
Ogievetsky, and Heer 2011) as the display engine in a Shiny GUI, using
custom message passing between server and client.

@@

A tour path is a smooth sequence of projection matrices, \(p\times d\),
that when combined with a matrix of n data points, \(n\times p\), and a
rendering method, produces a steady stream of \(d\)-dimensional views of
the data. Each tour is initialised with the \texttt{new\_tour()} method,
which instantiates a tour object and takes as arguments the data \(X\),
the tour method, e.g. \texttt{guided\_tour()}, and the starting basis.
Once initialised, a new target plane is chosen, and a series of steps
along a geodesic path from starting to target plane are generated by
interpolation.

@@

D3.js (Data-Driven Documents) (Bostock, Ogievetsky, and Heer 2011) is a
JavaScript library for manipulating documents based on data. The
advantages of D3 are similar to those provided by Shiny: namely, an
industry standard with rich array of powerful, easy to use methods and
widgets that can be displayed on a wide variety of devices, with a large
user base. D3 works on data objects in the JavaScript Object Notation
(JSON) format, which are then parsed and used to display customisable
data visualisations.

@@

The projections are calculated using the tour object in an
\texttt{observe()} environment, which repeatedly runs the code until a
reactive variable is changed, at which point it is ``invalidated'' and
then re-started. The projections are calculated using the following code
block:

@@

\address{%
Ursula Laa\\
Monash University\\
School of Physics and Astronomy\\
}
\href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}
